A Thee-Step Approach for Assessing Landscape Connectivity via Simulated Dispersal: African Wild Dog Case Study

### 2022-11-18 Chapter_1/04_Manuscript/Manuscript_2022-11-07.tex
Revisions done

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{scalerel}
\usepackage{tikz}
\usetikzlibrary{svg.path}

@@

% To be able to put an ORCID
\definecolor{orcidlogocol}{HTML}{A6CE39}
\tikzset{
  orcidlogo/.pic={
    \fill[orcidlogocol] svg{M256,128c0,70.7-57.3,128-128,128C57.3,256,0,198.7,0,128C0,57.3,57.3,0,128,0C198.7,0,256,57.3,256,128z};
    \fill[white] svg{M86.3,186.2H70.9V79.1h15.4v48.4V186.2z}
                 svg{M108.9,79.1h41.6c39.6,0,57,28.3,57,53.6c0,27.5-21.5,53.6-56.8,53.6h-41.8V79.1z M124.3,172.4h24.5c34.9,0,42.9-26.5,42.9-39.7c0-21.5-13.7-39.7-43.7-39.7h-23.7V172.4z}
                 svg{M88.7,56.8c0,5.5-4.5,10.1-10.1,10.1c-5.6,0-10.1-4.6-10.1-10.1c0-5.6,4.5-10.1,10.1-10.1C84.2,46.7,88.7,51.3,88.7,56.8z};
  }
}
\newcommand\orcid[1]{\href{https://orcid.org/#1}{\mbox{\scalerel*{
\begin{tikzpicture}[yscale=-1,transform shape]
  \pic{orcidlogo};
\end{tikzpicture}
}{|}}}}

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{A Three-Step Approach for Assessing Landscape Connectivity via Simulated
Dispersal: African Wild Dog Case Study}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \orcid{0000-0001-7477-2642} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538}
}

@@

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
\noindent \textbf{Context} \quad Dispersal of individuals contributes to
long-term population persistence, yet requires a sufficient degree of landscape
connectivity. To date, connectivity has mainly been investigated using
least-cost analysis and circuit theory, two methods that make assumptions that
are hardly applicable to dispersal. While these assumptions can be relaxed by
explicitly simulating dispersal trajectories across the landscape, a unified
approach for such simulations is lacking.

\noindent \textbf{Objectives} \quad Here, we propose and apply a simple
three-step approach to simulate dispersal and to assess connectivity using
empirical GPS movement data and a set of habitat covariates.

\noindent \textbf{Methods} \quad In step one of the proposed approach, we use
integrated step-selection functions to fit a mechanistic movement model
describing habitat and movement preferences of dispersing individuals. In step
two, we apply the parameterized model to simulate dispersal across the study
area. In step three, we derive three complementary connectivity maps; a heatmap
highlighting frequently traversed areas, a betweenness map pinpointing dispersal
corridors, and a map of inter-patch connectivity indicating the presence and
intensity of functional links between habitat patches. We demonstrate the
applicability of the proposed three-step approach in a case study in which we
use GPS data collected on dispersing African wild dogs (\textit{Lycaon pictus})
inhabiting northern Botswana.

\noindent \textbf{Results} \quad Using step-selection functions we successfully
parametrized a detailed dispersal model that described dispersing individuals'
habitat and movement preferences, as well as potential interactions among the
two. The model substantially outperformed a model that omitted such interactions
and enabled us to simulate 80,000 dispersal trajectories across the study area.

\noindent \textbf{Conclusion} \quad By explicitly simulating dispersal
trajectories, our approach not only requires fewer unrealistic assumptions about
dispersal, but also permits the calculation of multiple connectivity metrics
that together provide a comprehensive view of landscape connectivity. In our
case study, the three derived connectivity maps revealed several wild dog
dispersal hotspots and corridors across the extent of our study area. Each map
highlighted a different aspect of landscape connectivity, thus emphasizing their
complementary nature. Overall, our case study demonstrates that a
simulation-based approach offers a simple yet powerful alternative to
traditional connectivity modeling techniques. It is therefore useful for a
variety of applications in ecological, evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
% \subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.2000, Frankham.2002, Leigh.2012, Baguette.2013,
LaPoint.2013}, rescue non-viable populations \citep{Brown.1977}, and to colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the ability
to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Doerr.2011, Rudnick.2012}. Identifying dispersal corridors not only
necessitates a comprehensive understanding of the factors that limit dispersal,
but also an appropriate model to estimate connectivity \citep{Baguette.2013,
Vasudev.2015, Hofmann.2021}. To date, the most commonly used connectivity models
are least-cost path analysis (LCPA; \citealp{Adriaensen.2003}) and circuit
theory (CT; \citealp{McRae.2006, McRae.2008}). Unfortunately, both models rest
on assumptions that appear unsuitable for dispersers, thus calling for the
development of alternative approaches. One promising alternative is to assess
landscape connectivity via simulated dispersal trajectories generated from
individual-based movement models (IBMMs, \citealp{Diniz.2019}). However, IBMMs
require a large number of subjective modeling decisions, thus making
among-system comparisons difficult.

% Issues with Both Methods
% \subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, meaning that dispersal trajectories are usually strongly
directional. An interesting generalization that bridges the continuum between
LCPA and CT has been proposed by \citep{Panzacchi.2016} and enables to
capitalize on the merits of both approaches. Despite these and several other
generalizations of LCPA and CT (e.g. \citealp{Pinto.2009, Landguth.2012,
Panzacchi.2016, Brennan.2020}), some shortcomings remain. Most notably, all of
these methods rely on static permeability or resistance surfaces that can't
reflect the temporal dimension of dispersal. This permits statements about the
expected duration for moving between habitat patches \citep{Martensen.2017,
Diniz.2019}.

% What about IBMMs?
% \subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
using IBMMs and by converting simulated trajectories into meaningful measures of
connectivity \citep{Diniz.2019}. In contrast to LCPA and CT, IBMMs allow to
explicitly simulate how individuals move across and interact with the
encountered landscape \citep{Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020}, as well as to render potential interactions
between movement behavior and habitat conditions \citep{Avgar.2016}. This shifts
the focus towards a more functional view on connectivity
\citep{Tischendorf.2000}. Furthermore, IBMMs generate movement sequentially,
i.e. they generate a series of steps, so that the temporal dimension of
dispersal becomes explicit and allows modeling autocorrelation between
successive steps \citep{Diniz.2019}. Finally, simulations from IBMMs do not
enforce movement or connections towards preconceived endpoints but allow
individuals to adjust their route  ``on the go'', thereby preventing biases
arising from misplaced endpoints. Despite these advantages, a unifying approach
to simulate dispersal and assess connectivity using IBMMs is lacking.
Considering the large number of subjective decisions entailed by IBMMs, an
approach that streamlines and standardizes the application of dispersal
simulations to assess connectivity will, however, be critical to safeguard
comparability among studies.

@@

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis. First, GPS
    data and habitat covariates must be collected. The combined data is then
    analyzed using an integrated step selection model (step 1). The parametrized
    model is then treated as an individual-based movement model and used to
    simulate dispersal trajectories (step 2). Ultimately, simulated trajectories
    are translated into a set of maps that are pertinent to landscape
    connectivity (step 3). This includes a heatmap, indicating the traversal
    frequency across each spatial unit of the study area, a betweenness map,
    highlighting movement corridors and bottlenecks, and, finally, an
    inter-patch connectivity map, where the frequency of connections and their
    average duration can be depicted.}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

\section{Methods}
\subsection{Case Study}
\subsubsection{GPS Data}
We applied the three step approach presented in \Cref{GraphicalAbstract} to GPS
movement data from 16 dispersing African wild dog coalitions (7 female and 9
male coalitions). This data has been collected between 2011 and 2019 from a
free-ranging wild dog population in northern Botswana. During dispersal, GPS
collars recorded a fix every 4 hours and regularly transmitted data over the
Iridium satellite system. To ensure comparable time intervals between GPS fixes,
we removed any fixes that were not successfully obtained at the desired 4-hour
schedule (allowing for a tolerance of \( \pm \) 15 minutes). To prepare the data
for step-selection analysis, we converted the fixes (n = 4'169) into steps,
where each step represented the straight-line movement between two consecutive
GPS fixes \citep{Turchin.1998}. We only considered steps with equal
step-durations (i.e. 4 hours) for further analysis. We will refer to these steps
as ``realized steps''. We did not differentiate between sexes, for previous
research found little differences between sexes during dispersal
\citep{Woodroffe.2019, Cozzi.2020}. Additional details on the data collection
and preparation can be found in \cite{Cozzi.2020} and \cite{Hofmann.2021}.

@@

\subsubsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA,
\Cref{StudyArea}a and b) and encompassed a rectangular extent of roughly 1.3
Mio. km\textsuperscript{2}. With an area of 520'000 km\textsuperscript{2}, the
KAZA-TFCA is the world's largest transboundary conservation area and comprises
parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia, thus hosting a rich
diversity of landscapes, ranging from savannah to grassland and from dry to
moist woodland habitats. In its center lies the Okavango Delta, a dominant
hydro-geographical feature and the world's largest flood-pulsing inland delta.
Large portions of the KAZA-TFCA are formally protected in the form of national
parks (NPs) or other protected areas, yet a considerable portion of the
landscape remains human-dominated (e.g. roads, agricultural sites, and
settlements).

@@

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) to each step (\ref{EQ1};
\citealp{Fortin.2005}), where \(w(x)\) depended on the step's associated
covariates (\(x_1, x_2, ..., x_n\)) and on the animal's relative selection
strengths \citep{Avgar.2017} towards these covariates (\(\beta_1, \beta_2,
..., \beta_n\)):

@@

Our movement model was based on a habitat selection model that was previously
developed for dispersing wild dogs (hereafter referred to as \textit{base
model}, \citealp{Hofmann.2021}). In the base model, no interactions among
habitat covariates and movement covariates were considered, so we here expanded
the model and allowed for such interactions, acknowledging that movement
preferences during dispersal could depend on habitat conditions (details in
Appendix A1). To determine the most parsimonious movement model among model
candidates, we ran stepwise forward model selection based on Akaike's
Information Criterion (AIC, \citealp{Burnham.2002}). More specifically, we
started with the base model and iteratively increased model complexity by adding
all possible interactions between movement and habitat covariates. Given that
the focus of our analysis lied on predicting dispersal patterns and all model
candidates were biologically intuitive, we deemed the use of model selection
appropriate. However, caution should be employed if causal relationships are of
interest, as model selection may lead to biased parameter estimate
\citep{Whittingham.2006}. We validated the predictive power of the most
parsimonious model using k-fold cross-validation for case-control studies as
described in \cite{Fortin.2009}. This validation attests significant prediction
ability to the movement model if the model outperforms a random guess and
systematically assigns low ranks (high selection scores) to observed steps
(details in Appendix A2).

@@

\subsubsection{Betweenness Map}
To pinpoint movement corridors and bottlenecks, we converted simulated
trajectories into a network and calculated betweenness scores for all
raster-cells in the study area \citep{BastilleRousseau.2018}. Betweenness is a
pertinent metric for connectivity as it measures how often a specific
network-node (in our case a raster-cell) lies on a shortest path between any
other pair of nodes \citep{BastilleRousseau.2018}. To convert simulated
trajectories into a network, we followed \cite{BastilleRousseau.2018} and
overlaid the study area (including the buffer) with a raster containing 2.5 km x
2.5 km raster-cells, where the center of each raster-cell served as node in the
final network. To identify edges (i.e. connections) between the nodes, we used
the simulated trajectories and determined all transitions occurring from one
cell to another, as well as the frequency at which those transitions occurred
(see also Appendix A4). This resulted in an edge-list that we translated into a
weighted network using the r-package {\tt igraph} \citep{Gabor.2006}. The final
weight of each edge was determined by the frequency of transitions, yet because
{\tt igraph} handles edge weights (\(\omega\)) as costs, we inverted the
traversal-frequency through each raster-cell by applying \(\omega =
\frac{mean(Traversal Frequency)}{Traversal Frequency_i}\). Consequently,
regularly used edges received small weights (i.e. low costs) and vice versa. We
used the weighted network to calculate betweenness scores for all network nodes.

@@

\subsubsection{Validation}
To validate our predictions of connectivity, we utilized additional dispersal
data that was collected on eight dispersing coalitions between the years 2019
and 2022 (totalling to 2'668 GPS locations). We used path selection analysis
(PSF, \citealp{Cushman.2010}) to assess if observed dispersal trajectories
followed areas of high predicted connectivity. Similar to SSF, PSF enables to
detect selection for certain features by comparing observed paths to randomly
generated paths. Here, we paired each observed path with 50 random paths that we
generated by randomly rotating and shifting observed paths by a random angle
\(\alpha \sim U(-\pi, +\pi)\) and a random distance \(d \sim U(0 \) km, \(50\)
km\(\)). Along each path, we then extracted connectivity values from the heatmap
(see above) generated after 68, 125, 250, 500, and 2'000 simulated steps,
respectively. Finally, we ran conditional logistic regression to contrast
observed and random paths. In case of systematic selection for high-connectivity
areas, the regression coefficients from the corresponding conditional logistic
regression model should be positive.

\section{Results}
% \subsection{Movement Model}
The most parsimonious movement model consisted of movement covariates, habitat
covariates, as well as several of their interactions, thus suggesting that
movement behavior during dispersal depended on habitat conditions
(\Cref{MovementModel}a, Table S1 and Table S2). Although multiple models
received an AIC weight \(>\) 0 (Table S1), we only considered results from the
most parsimonious model for simplicity. This decision only marginally influenced
subsequent steps as all models with positive AIC weights retained similar
covariates (Table S1). The k-fold cross-validation showed that the final model
substantially outperformed a random guess and provided reliable predictions
(i.e. confidence intervals of \(\bar{r}_{s, realized}\) and \(\bar{r}_{s,
random}\) did not overlap). Moreover, the model correctly assigned high
selection scores to realized steps (\Cref{MovementModel}b), indicating a good
fit between predictions and observations. As can be taken from the Spearman rank
correlation coefficient, the inclusion of several interactions between movement
and habitat covariates significantly improved model performance (\(\bar{r}_{s,
realized} = -0.65; 95\%-CI = [-0.67, -0.64]\)), compared to the base model
(\(\bar{r}_{s, realized} = -0.55; 95\%-CI = [-0.57, -0.52]\);
\citealp{Hofmann.2021}). Our validation of the resulting connectivity maps using
independent dispersal data showed that dispersers preferentially followed areas
of high predicted connectivity, as coefficients from the PSF models were all
significantly greater than zero (\Cref{MovementModel}c). The movement model thus
successfully predicted functional connectivity.

Plots that aid with the interpretation of the most parsimonious movement model
are provided in Figure S3 and suggest that, under average conditions, dispersing
wild dogs avoided moving through water, woodlands, and areas dominated by
humans, but preferred moving across shrublands or grasslands
(\Cref{MovementModel}a). Dispersers realized shorter steps (indicating slower
movements) in areas covered by water or woodland, while realizing larger steps
in areas dominated by shrubs or grass (\Cref{MovementModel}a). We found a
particularly large effect for the variable \textsf{LowActivity}, suggesting that
dispersing wild dogs moved substantially faster during twilight and at night
(i.e. between 17:00 and 09:00 o'clock; \Cref{MovementModel}a). Although
dispersers revealed a preference for directional movements (i.e. low turning
angles), especially when moving quickly, they did less so in proximity to humans
or water, resulting in more tortuous movements in such areas
(\Cref{MovementModel}a).

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage[colorinlistoftodos]{todonotes}

@@

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{A Thee-Step Approach for Assessing Landscape Connectivity via Simulated
Dispersal: African Wild Dog Case Study}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \and
  Gabriele Cozzi\textsuperscript{1,2} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \and
  Dominik M. Behr\textsuperscript{1,2}
}

@@

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
\noindent 1. Dispersal of individuals contributes to long-term population
persistence, yet requires a sufficient degree of landscape connectivity. To
date, connectivity has mainly been investigated using least-cost analysis and
circuit theory, two methods that make assumptions that are hardly applicable to
dispersal. While these assumptions can be relaxed by explicitly simulating
dispersal trajectories across the landscape, a unified approach for such
simulations is lacking.

\noindent 2. Here, we present a simple three-step approach to simulate dispersal
and to assess connectivity using empirical GPS movement data and a set of
habitat covariates. In step one, we use integrated step-selection functions to
fit a mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, we apply the parameterized model to
simulate dispersal across the study area. In step three, we derive three
complementary connectivity maps; a heatmap highlighting frequently traversed
areas, a betweenness map pinpointing dispersal corridors, and a map of
inter-patch connectivity indicating the presence and intensity of functional
links between habitat patches. We demonstrate the applicability of the proposed
three-step approach in a case study in which we use GPS data collected on
dispersing African wild dogs (\textit{Lycaon pictus}) inhabiting northern
Botswana.

\noindent 3. Using step-selection functions we successfully parametrized a
detailed dispersal model that described dispersing individuals habitat and
movement preferences, as well as potential interactions among the two. The model
substantially outperformed a model that omitted such interactions and enabled us
to simulate 80,000 dispersal trajectories across the study area.

\noindent 4. By explicitly simulating individuals, our approach not only
requires fewer unrealistic assumptions about dispersal, but also permits the
calculation of multiple connectivity metrics that together provide a
comprehensive view of landscape connectivity. In our case study, the three
derived connectivity maps revealed several wild dog dispersal hotspots and
corridors across the extent of our study area. Each map highlighted a different
aspect of landscape connectivity, thus emphasizing their complementary nature.
Overall, our case study demonstrates that a simulation-based approach offers a
simple yet powerful alternative to traditional connectivity modeling techniques.
It is therefore useful for a variety of applications in ecological,
evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
% \subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.2000, Frankham.2002, Leigh.2012, Baguette.2013},
rescue non-viable populations \citep{Brown.1977}, and to colonize unoccupied
habitats \citep{Hanski.1999b, MacArthur.2001}. However, the ability to disperse
depends on a sufficient degree of landscape connectivity \citep{Fahrig.2003,
Clobert.2012}, making the identification and protection of dispersal corridors
that promote connectivity a task of fundamental importance \citep{Doerr.2011,
Rudnick.2012}. Identifying dispersal corridors not only necessitates a
comprehensive understanding of the factors that limit dispersal, but also an
appropriate model to estimate connectivity \citep{Baguette.2013, Vasudev.2015,
Hofmann.2021}. To date, the most commonly used connectivity models are
least-cost path analysis (LCPA; \citealp{Adriaensen.2003}) and circuit theory
(CT; \citealp{McRae.2006, McRae.2008}). Unfortunately, both models rest on
assumptions that appear unsuitable for dispersers, thus calling for the
development of alternative approaches. One promising alternative is to assess
landscape connectivity via simulated dispersal trajectories generated from
individual-based movement models (IBMMs, \citealp{Diniz.2019}). However, IBMMs
require a large number of subjective modeling decisions, thus making
among-system comparisons difficult.

% Issues with Both Methods
% \subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, meaning that dispersal trajectories are usually strongly
directional. Moreover, because both models require static permeability or
resistance surfaces as input, they are unable to reflect the temporal dimension
of dispersal and therefore fail to allow statements about the expected duration
for moving between habitat patches \citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
% \subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
using IBMMs and by converting simulated trajectories into meaningful measures of
connectivity \citep{Diniz.2019}. In contrast to LCPA and CT, IBMMs allow to
explicitly simulate how individuals move across and interact with the
encountered landscape \citep{Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020}, as well as to render potential interactions
between movement behavior and habitat conditions \citep{Avgar.2016}. This
strictly shifts the focus from a structural to a more functional view on
connectivity \citep{Tischendorf.2000}. Furthermore, IBMMs generate movement
sequentially, i.e. they generate a series of steps, so that the temporal
dimension of dispersal becomes explicit and allows modeling autocorrelation
between successive steps \citep{Diniz.2019}. Finally, simulations from IBMMs do
not enforce movement or connections towards preconceived endpoints but allow
individuals to adjust their route  ``on the go'', thereby preventing biases
arising from misplaced endpoints. Despite these advantages, a unifying approach
to simulate dispersal and assess connectivity using IBMMs is lacking.
Considering the large number of subjective decisions entailed by IBMMs, an
approach that streamlines and standardizes the application of dispersal
simulations to assess connectivity will, however, be critical to safeguard
comparability among studies.

@@

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis. First, GPS
    data and habitat covariates must be collected. The combined data is then
    analyzed using an integrated step selection model. The parametrized model is
    then treated as an individual-based movement model and used to simulate
    dispersal trajectories. Ultimately, simulated trajectories are translated
    into a set of maps that are pertinent to landscape connectivity. This
    includes a heatmap, indicating the traversal frequency across each spatial
    unit of the study area, a betweenness map, highlighting movement corridors
    and bottlenecks, and, finally, an inter-patch connectivity map, where the
    frequency of connections and their average duration can be depicted.}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

\section{Methods}
\subsection{Case Study}
\subsubsection{GPS Data}
We applied the three step approach presented in \Cref{GraphicalAbstract} to GPS
movement data from 16 dispersing African wild dog coalitions (7 female and 9
male coalitions). This data has been collected between 2011 and 2019 from a
free-ranging wild dog population in northern Botswana. During dispersal, GPS
collars recorded a fix every 4 hours and regularly transmitted data over the
Iridium satellite system. To ensure comparable time intervals between GPS fixes,
we removed any fixes that were not successfully obtained at the desired 4-hour
schedule (allowing for a tolerance of \( \pm \) 15 minutes). To prepare the data
for step-selection analysis, we converted the fixes (n = 4'169) into steps,
where each step represented the straight-line movement between two consecutive
GPS fixes \citep{Turchin.1998}. We only considered steps with equal
step-durations (i.e. 4 hours) for further analysis. We will refer to these steps
as ``realized steps''. Additional details on the data collection and preparation
can be found in \cite{Cozzi.2020} and \cite{Hofmann.2021}.

@@

\subsubsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA,
\Cref{StudyArea}a and b). The KAZA-TFCA is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, thus hosting a rich diversity of landscapes, ranging from savannah
to grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydro-geographical feature and the world's largest
flood-pulsing inland delta. Large portions of the KAZA-TFCA are formally
protected in the form of national parks (NPs) or other protected areas, yet a
considerable portion of the landscape remains human-dominated (e.g. roads,
agricultural sites, and settlements).

@@

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) to each step (\ref{EQ1};
\citealp{Fortin.2005}), where \(w(x)\) depended on the step's associated
covariates (\(x_1, x_2, ..., x_n\)) and on the animal's relative selection
strengths \citealp{Avgar.2017} towards these covariates (\(\beta_1, \beta_2,
..., \beta_n\)):

@@

Our movement model was based on a habitat selection model that was previously
developed for dispersing wild dogs (hereafter referred to as \textit{base
model}, \citealp{Hofmann.2021}). In the base model, no interactions among
habitat covariates and movement covariates were considered, so we here expanded
the model and allowed for such interactions, acknowledging that movement
preferences during dispersal could depend on habitat conditions (details in
Appendix A1). To determine the most parsimonious movement model among model
candidates, we ran stepwise forward model selection based on Akaike's
Information Criterion (AIC, \citealp{Burnham.2002}). More specifically, we
started with the base model and iteratively increased model complexity by adding
all possible interactions between movement and habitat covariates. We validated
the predictive power of the most parsimonious model using k-fold
cross-validation for case-control studies as described in \cite{Fortin.2009}.
This validation proves a significant prediction in case the Spearman rank
correlation coefficient of predicted step-ranks and associated frequencies under
the movement model is significantly greater than under the assumption of random
preferences (details in Appendix A2).

@@

\subsubsection{Betweenness Map}
To pinpoint movement corridors and bottlenecks, we converted simulated
trajectories into a network and calculated betweenness scores for all
raster-cells in the study area \citep{BastilleRousseau.2018}. Betweenness is a
pertinent metric for connectivity as it measures how often a specific
network-node (in our case a raster-cell) lies on a shortest path between any
other pair of nodes \citep{BastilleRousseau.2018}. To convert simulated
trajectories into a network, we followed \cite{BastilleRousseau.2018} and
overlaid the study area (including the buffer) with a raster containing 5 km x 5
km raster-cells, where the center of each raster-cell served as node in the
final network. To identify edges (i.e. connections) between the nodes, we used
the simulated trajectories and determined all transitions occurring from one
cell to another, as well as the frequency at which those transitions occurred.
This resulted in an edge-list that we translated into a weighted network using
the r-package {\tt igraph} \citep{Gabor.2006}. The final weight of each edge was
determined by the frequency of transitions, yet because {\tt igraph} handles
edge weights (\(\omega\)) as costs, we inverted the traversal-frequency through
each raster-cell by applying \(\omega = \frac{mean(Traversal
Frequency)}{Traversal Frequency_i}\). Consequently, regularly used edges
received small weights (i.e. low costs) and vice versa. We used the weighted
network to calculate betweenness scores for all network nodes.

@@

\section{Results}
% \subsection{Movement Model}
The most parsimonious movement model consisted of movement covariates, habitat
covariates, as well as several of their interactions, thus suggesting that
movement behavior during dispersal depended on habitat conditions
(\Cref{MovementModel}a, Table S1 and Table S2). Although multiple models
received an AIC weight \(>\) 0 (Table S1), we only considered results from the
most parsimonious model for simplicity. This decision only marginally influenced
subsequent steps as all models with positive AIC weights retained similar
covariates (Table S1). The k-fold cross-validation showed that the final model
substantially outperformed a random guess and provided reliable predictions
(i.e. confidence intervals of \(\bar{r}_{s, realized}\) and \(\bar{r}_{s,
random}\) did not overlap). Moreover, the model correctly assigned high
selection scores to realized steps (\Cref{MovementModel}b), indicating a good
fit between predictions and observations. As can be taken from the Spearman rank
correlation coefficient, the inclusion of several interactions between movement
and habitat covariates significantly improved model performance (\(\bar{r}_{s,
realized} = -0.65; 95\%-CI = [-0.67, -0.64]\)), compared to the base model
(\(\bar{r}_{s, realized} = -0.55; 95\%-CI = [-0.57, -0.52]\);
\citealp{Hofmann.2021})

Plots that aid with the interpretation of the most parsimonious movement model
are provided in Figure S2 and suggest that, under average conditions, dispersing
wild dogs avoided moving through water, woodlands, and areas dominated by
humans, but preferred moving across shrublands or grasslands
(\Cref{MovementModel}a). Dispersers realized shorter steps (indicating slower
movements) in areas covered by water or woodland, while realizing larger steps
in areas dominated by shrubs or grass (\Cref{MovementModel}a). We found a
particularly large effect for the variable \textsf{LowActivity}, suggesting that
dispersing wild dogs moved substantially faster during twilight and at night
(i.e. between 17:00 and 09:00 o'clock; \Cref{MovementModel}a). Although
dispersers revealed a preference for directional movements (i.e. low turning
angles), especially when moving quickly, they did less so in proximity to humans
or water, resulting in more tortuous movements in such areas
(\Cref{MovementModel}a).

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{99_MovementModel}
    \caption{(a) Most parsimonious movement model for dispersing wild dogs. The
    model comprises a habitat kernel, a movement kernel, as well as their
    interactions. The horizontal line segments delineate the 90\%, 95\%, and
    99\% confidence-intervals for the respective \(\beta\)-coefficients.
    Significance codes: * \(p < 0.10\), ** \(p < 0.05\), *** \(p < 0.01\). (b)
    Results from the k-fold cross validation procedure. The upper plot shows
    rank frequencies of realized steps according to model predictions with known
    preferences, whereas the lower plot shows rank frequencies of realized steps
    when assuming random preferences. The blue ribbon shows the prediction
    interval around a loess smoothing regression that we fitted to ease the
    interpretation of the plots. The significant correlation between rank and
    associated frequency in (b1) highlights that the most parsimonious model
    successfully outperformed a random guess (b2) and assigned high selection
    scores to realized steps.}
    \label{MovementModel}
  \end{center}
\end{figure}

\subsection{Dispersal Simulation}
Dispersal simulations based on the most parsimonious movement model proved
useful for assessing landscape connectivity. Of the 50,000 simulated dispersal
trajectories that originated from the main study area, only 4.5\% reached a map
boundary, suggesting that we successfully mitigated biases from boundary
effects. Moreover, our examination of the relative traversal frequency across
all checkpoints showed that the relative traversal frequency reached a steady
state after 10,500 simulated dispersal trajectories (Figure S3). Although
variability in the relative traversal frequency kept decreasing as we increased
the number of simulated dispersers, the marginal benefit of simulating
additional trajectories diminished quickly (Figure S3).
